---
title: "Reproducible example of current open issues with the SpatialKWD package"
output:
  html_document:
    df_print: paged
    toc: true
---

2023-08-03

## 1. Preparing example data

```{r}
library(SpatialKWD)
library(sp)
library(raster)
library(sdcSpatial)
```

The issues are illustrated by means of an example data set from the sdcSpatial package called
'dwellings'. The first step is to create the unprotected raster map, which forms the
ground truth. 

The example uses a map of counts (no. of units per grid cell), with 200m $\times$ 
200m grid cells in a 62 $\times$ 61 cells raster map.

```{r}
data("dwellings")

dw <- sdcSpatial::sdc_raster(dwellings[, c("x", "y")], 
                             variable = 1, r = 200,
                             max_risk = .8, min_count = 10)

ras_orig <- dw$value$count # original raster

raster::plot(ras_orig, main = "no. of dwellings") # map
ras_orig # raster info
```

Next, we create the altered raster maps using three protection methods:
1) suppression of sensitive cells (remove_sensitive),
2) quadtree method (protect_quadtree),
3) kernel smoothing (protect_smooth).

```{r}
# apply protection methods
dw_rm <- remove_sensitive(dw)
dw_qt <- protect_quadtree(dw, max_zoom = 2)
dw_sm <- protect_smooth(dw, bw = 200)

# extract protected rasters from sdcSpatial object
ras_rm <- dw_rm$value$count
ras_qt <- dw_qt$value$count
ras_sm <- dw_sm$value$count
```

Following up, we select a square focus area. It is identified by its centroid and 
radius.

```{r}
# Focus area (square)

fa_ctrd <- c(155100, 464100) # center of focus area
fa_rads <- 3000              # radius of focus area


# maps

par(mfrow = c(2, 2))

plot(ras_orig, main = "original")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")
plot(ras_rm, main = "cell suppression")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")
plot(ras_qt, main = "quad tree method")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")
plot(ras_sm, main = "smoothing method")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")

par(mfrow = c(1, 1))
```


## 2. FocusArea sometimes returns zero for clear differences

We continue by evaluating the difference between altered maps and ground truth with 
respect to the chosen focus area.

This involves the following steps:
1) extract the center coordinates of tiles (raster::xyFromCell),
2) extract the tile-values (weights) (raster::getValues) [where NA-values are
treated as zeros].

```{r}
# extract coordinates from raster
xy <- raster::xyFromCell(ras_orig, 1:ncell(ras_orig))

# extract cell values from rasters
w_orig <- raster::getValues(ras_orig)
w_rm   <- raster::getValues(ras_rm)
w_qt   <- raster::getValues(ras_qt)
w_sm   <- raster::getValues(ras_sm)

# treat NA-nodes as having zero weight
w_orig[is.na(w_orig)] <- 0
w_rm[is.na(w_rm)] <- 0
w_qt[is.na(w_qt)] <- 0
w_sm[is.na(w_sm)] <- 0
```

Using these inputs we calculate the first set of KWDs.

```{r}
kwd_rm <- SpatialKWD::focusArea(Coordinates = xy, Weights = cbind(w_orig, w_rm),
                                L = 3, recode = TRUE, area = "linf",
                                x = fa_ctrd[1], y = fa_ctrd[2], radius = fa_rads)
kwd_qt <- SpatialKWD::focusArea(Coordinates = xy, Weights = cbind(w_orig, w_qt),
                                L = 3, recode = TRUE, area = "linf",
                                x = fa_ctrd[1], y = fa_ctrd[2], radius = fa_rads)
kwd_sm <- SpatialKWD::focusArea(Coordinates = xy, Weights = cbind(w_orig, w_sm),
                                L = 3, recode = TRUE, area = "linf",
                                x = fa_ctrd[1], y = fa_ctrd[2], radius = fa_rads)
```

Even though the suppression map is clearly different (all cells with < 10 units are 
removed, see maps above), we get a KWD of zero.

```{r}
c(kwd_rm$distance, kwd_qt$distance, kwd_sm$distance)
```

Second try: The description of the focusArea function tells us that "All the weights outside 
the focus area should be equal to zero." We create new rasters, where all tiles outside
the focus area are assigned zero values.

```{r}
ex_fa <- extent(fa_ctrd[1] - (fa_rads - 100), fa_ctrd[1] + (fa_rads - 100),
                fa_ctrd[2] - (fa_rads - 100), fa_ctrd[2] + (fa_rads - 100))
clls_fa <- cellsFromExtent(ras_orig, ex_fa)

ras_rm_fa <- ras_rm
ras_qt_fa <- ras_qt
ras_sm_fa <- ras_sm
ras_rm_fa[-clls_fa] <- 0
ras_qt_fa[-clls_fa] <- 0
ras_sm_fa[-clls_fa] <- 0


par(mfrow = c(2, 2))

plot(ras_orig, main = "original")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")
plot(ras_rm_fa, main = "cell suppression (focus area)")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")
plot(ras_qt_fa, main = "quad tree method (focus area)")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")
plot(ras_sm_fa, main = "smoothing method (focus area)")
rect(xleft = fa_ctrd[1] - fa_rads, xright = fa_ctrd[1] + fa_rads,
     ybottom = fa_ctrd[2] - fa_rads, ytop = fa_ctrd[2] + fa_rads, border = "red")

par(mfrow = c(1, 1))
```

With these we re-calculate the KWDs.

```{r}
# updated weights (zero outside focus area)
w_rm_fa <- getValues(ras_rm_fa)
w_qt_fa <- getValues(ras_qt_fa)
w_sm_fa <- getValues(ras_sm_fa)
w_rm_fa[is.na(w_rm_fa)] <- 0
w_qt_fa[is.na(w_qt_fa)] <- 0
w_sm_fa[is.na(w_sm_fa)] <- 0

kwd_rm_fa <- SpatialKWD::focusArea(Coordinates = xy, Weights = cbind(w_orig, w_rm_fa),
                                   L = 3, recode = TRUE, area = "linf",
                                   x = fa_ctrd[1], y = fa_ctrd[2], radius = fa_rads)
kwd_qt_fa <- SpatialKWD::focusArea(Coordinates = xy, Weights = cbind(w_orig, w_qt_fa),
                                   L = 3, recode = TRUE, area = "linf",
                                   x = fa_ctrd[1], y = fa_ctrd[2], radius = fa_rads)
kwd_sm_fa <- SpatialKWD::focusArea(Coordinates = xy, Weights = cbind(w_orig, w_sm_fa),
                                   L = 3, recode = TRUE, area = "linf",
                                   x = fa_ctrd[1], y = fa_ctrd[2], radius = fa_rads)
```

We still get a KWD of zero for the 'removal' method. Furthermore, we now also get zero 
for smoothing.

```{r}
c(kwd_rm_fa$distance, kwd_qt_fa$distance, kwd_sm_fa$distance)
```


## 3. How to apply a fixed cost to missing units of mass?

Suppose now we look at the whole map rather than focus areas. Then, we might get 
a mass mismatch, especially with removal:

```{r}
c(sum(w_orig), sum(w_rm))
```

There are ca. 2000 units missing after removal. We want to try solving the problem
by assigning each a cost of half the map diagonal.

```{r}
# compute length of the diagonal in cells
diag_cells <- sqrt(ncol(ras_orig)^2 + nrow(ras_orig)^2)
# ... in meters
diag_metrs <- diag_cells * 200

# solve with extra artificial bin (?)
kwd_rm_bal <- compareOneToOne(Coordinates = xy, Weights = cbind(w_orig, w_rm),
                              L = 3, recode = TRUE,
                              unbalanced = TRUE, unbal_cost = diag_metrs/2)
```

Again, this yields a distance of zero.

```{r}
kwd_rm_bal$distance
```

Do we get a distance of zero, because all non-zero tiles after removal match exactly? 
(I.e. for any given tile the mass is missing completely or not at all.)

Given that this is the case, we could, in principle, assign the costs manually 
(2000 $\times c$ ).
Still, intuitively the function should not return zero (?)

